#Machine Learning Course Project
*by Venkatesh*

## Synopsis
The goal of this analysis to predict the manner in which the six participants did the exercise. They perform barbell lifts correctly and incorrectly in 5 different ways. 

Given both training and test data from the following study:

*Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. Qualitative Activity Recognition of Weight Lifting Exercises. Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13) . Stuttgart, Germany: ACM SIGCHI, 2013.*

The prediction model is to be run on the testing data to predict the outcome of 20 different test cases.
Loading all the required packages for the analysis

```{r, echo=TRUE}
library(AppliedPredictiveModeling)
library(caret)
library(rattle)
library(rpart.plot)
library(randomForest)
library(corrplot)
```

## Question
In the aforementioned study, six participants participated in a dumbell lifting exercise five different ways. The five ways, as described in the study, were “exactly according to the specification (Class A), throwing the elbows to the front (Class B), lifting the dumbbell only halfway (Class C), lowering the dumbbell only halfway (Class D) and throwing the hips to the front (Class E). Class A corresponds to the specified execution of the exercise, while the other 4 classes correspond to common mistakes.”

##Input Data

The first step is to import the data and to verify that the training data and the test data are identical.
By processing data gathered from accelerometers on the belt, forearm, arm, and dumbell of the participants in a machine learning algorithm, the question is can the appropriate activity quality (class A-E) be predicted?
```{r, echo=TRUE}
# Download data.
url_raw_training <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
file_dest_training <- "pml-training.csv"

url_raw_testing <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
file_dest_testing <- "pml-testing.csv"

# Import the data treating empty values as NA.
df_training <- read.csv(file_dest_training, na.strings=c("NA",""), header=TRUE)
colnames_train <- colnames(df_training)
df_testing <- read.csv(file_dest_testing, na.strings=c("NA",""), header=TRUE)
colnames_test <- colnames(df_testing)

# Verify that the column names (excluding classe and problem_id) are identical in the training and test set.
all.equal(colnames_train[1:length(colnames_train)-1], colnames_test[1:length(colnames_train)-1])
```

##Features
We have training and testing data sets. The training data have 19622 records with 160 variables, while the test data set has 20 records. Some variables have NA for almost all records such as kurtosis_roll_belt and kurtosis_picth_belt. kurtosis_picth_belt has 19,248 NA out of 19,622 records. The problem happens to variables whose name start with max, min, kurtosis, skewness, var, avg and apmlitude. These variables will not be useful for the classification even if we impute the data, because almost all records will have identical values. Column X shows the row number, column user_name is a name of the participants, and columns raw_timestamp_part_1, raw_timestamp_part_2 and cvtd_timestamp mean date time, which are not helpful for the prediction. They were removed from the analysis.
```{r, echo=TRUE}
# Count the number of non-NAs in each col.
nonNAs <- function(x) {
    as.vector(apply(x, 2, function(x) length(which(!is.na(x)))))
}

# Build vector of missing data or NA columns to drop.
colcnts <- nonNAs(df_training)
drops <- c()
for (cnt in 1:length(colcnts)) {
    if (colcnts[cnt] < nrow(df_training)) {
        drops <- c(drops, colnames_train[cnt])
    }
}

# Drop NA data and the first 7 columns as they're unnecessary for predicting.
df_training <- df_training[,!(names(df_training) %in% drops)]
df_training <- df_training[,8:length(colnames(df_training))]

df_testing <- df_testing[,!(names(df_testing) %in% drops)]
df_testing <- df_testing[,8:length(colnames(df_testing))]

# Show remaining columns.
colnames(df_training)
```

```{r, echo=TRUE}
colnames(df_testing)
```

We now split the updated training dataset into a training dataset (70% of the observations) and a validation dataset (30% of the observations). This validation dataset will allow us to perform cross validation when developing our model.
```{r, echo=TRUE}
inTrain = createDataPartition(y = df_training$classe, p = 0.7, list = FALSE)
small_train = df_training[inTrain, ]
small_valid = df_training[-inTrain, ]
```

At this point, our dataset contains 54 variables, with the last column containing the 'classe' variable we are trying to predict. We begin by looking at the correlations between the variables in our dataset. We may want to remove highly correlated predictors from our analysis and replace them with weighted combinations of predictors. This may allow a more complete capture of the information available.

```{r, echo=TRUE}
corMat <- cor(small_train[, -53])
corrplot(corMat, order = "FPC", method = "color", type = "lower", tl.cex = 0.5, tl.col = rgb(0, 0, 0))
```

This grid shows the correlation between pairs of the predictors in our dataset. From a high-level perspective darker blue and darker red squares indicate high positive and high negative correlations, respectively. Based on this observation, we choose to implement a principal components analysis to produce a set of linearly uncorrelated variables to use as our predictors.

##Principal Components Analysis and Machine Learning

We pre-process our data using a principal component analysis, leaving out the last column ('classe'). After pre-processing, we use the 'predict' function to apply the pre-processing to both the training and validation subsets of the original larger 'training' dataset.
```{r, echo=TRUE}
preProc <- preProcess(small_train[, -53], method = "pca", thresh = 0.99)
trainPC <- predict(preProc, small_train[, -53])
valid_testPC <- predict(preProc, small_valid[, -53])
```

Next, we train a model using a random forest approach on the smaller training dataset. We chose to specify the use of a cross validation method when applying the random forest routine in the 'trainControl()' parameter. Without specifying this, the default method (bootstrapping) would have been used. The bootstrapping method seemed to take a lot longer to complete, while essentially producing the same level of 'accuracy'.
```{r, echo=TRUE}
modelFit <- train(small_train$classe ~ ., method = "rf", data = trainPC, trControl = trainControl(method = "cv", 
    number = 4), importance = TRUE)
```

We now review the relative importance of the resulting principal components of the trained model, 'modelFit'.
```{r, echo=TRUE}
varImpPlot(modelFit$finalModel, sort = TRUE, type = 1, pch = 16, col = 1, cex = 1, 
    main = "Importance of the Individual Principal Components")
```

As you look from the top to the bottom on the y-axis, this plot shows each of the principal components in order from most important to least important. The degree of importance is shown on the x-axis–increasing from left to right. Therefore, points high and to the right on this graph correspond to those principal components that are especially valuable in terms of being able to classify the observed training data.

##Cross Validation Testing and Out-of-Sample Error Estimate

Call the 'predict' function again so that our trained model can be applied to our cross validation test dataset. We can then view the resulting table in the 'confusionMatrix' function's output to see how well the model predicted/classified the values in the validation test set (i.e. the 'reference' values)

```{r, echo=TRUE}
pred_valid_rf <- predict(modelFit, valid_testPC)
confus <- confusionMatrix(small_valid$classe, pred_valid_rf)
confus$table
```

The estimated out-of-sample error is 1.000 minus the model's accuracy, the later of which is provided in the output of the confusionmatrix, or more directly via the 'postresample' function.

```{r, echo=TRUE}
accur <- postResample(small_valid$classe, pred_valid_rf)
model_accuracy <- accur[[1]]
model_accuracy
```

```{r, echo=TRUE}
out_of_sample_error <- 1 - model_accuracy
out_of_sample_error
```

*The estimated accuracy of the model is 98.4% and the estimated out-of-sample error based on our fitted model applied to the cross validation dataset is 1.6%.*

##Predicted Results
Finally, we apply the pre-processing to the original testing dataset, after removing the extraneous column labeled 'problem_id' (column 54). We then run our model against the testing dataset and display the predicted results.
```{r, echo=TRUE}
testPC <- predict(preProc, df_testing[, -53])
pred_final <- predict(modelFit, testPC)
pred_final
```
